---
title: "Regression Models - Course Project"
author: "Gildardo Rojas Nandayapa"
date: "Saturday, October 25, 2014"
output:
  pdf_document:
    fig_height: 4
    keep_tex: yes
  html_document: default
---
## Executive Summary

Motor Trend, a leading magazine about the automobile industry is interested in looking at a data set of a collection of cars, exploring the relationship between a set of variables and miles per gallon (MPG). There is particular interest in the following two questions:

1. Is an automatic or manual transmission better for MPG?
2. What is the difference in MPG between automatic and manual transmissions?

Using linear regression analysis, we can determine that there is a signficant difference between the mean MPG for automatic and manual transmission cars. Manual transmissions achieve a higher value of MPG compared to automatic transmission.

Transmission type is relevant, but not an unique factor predicting the mpg outcome as we may see in the following analysis.

##The Data

The data was extracted from the 1974 Motor Trend US magazine, it comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973-74 models).

##Exploratory Analysis

This study is focused on the effects of car transmission type over mpg efficiency, so a simple box plot may help to depict the difference between cars with automatic and manual transmission. The plot shows that manual transmissions have higher mpg. See **Appendix - Figure 1. MPG by Transmission Type - Box plot**

To find relationships between all variables, data exploration can be extended visually using the pairs plot. By inspecting the plot we may notice variables related to mpg, though the impact varies in magnitude and slope. See **Appendix - Figure 2. Motor Trend Car Road Tests - Variable pairs plot.**

###Statistical Inference

**T-test**
Assuming a normal distribution, by performing a t-test we may observe that there is a significant difference between manual and automatic transmission types in the resulting mpg.

```{r}
data(mtcars)
t.test(mpg ~ am, data = mtcars)
```

Here we may also observe that the difference in mpg for vehicles with manual over automatic transmission type is 7.24.

##Regression Analysis

###Linear Regression

First we attempt simple linear regression.
```{r}
fit <- lm(mpg ~ am, data=mtcars)
summary(fit)
```

Observing the summary information, we can confirm that manual transmission vehicles have 7.24 MPG more than automatic ones. But this model doesn't provide complete information as the Multiple R-squared value of .3598 means the model can just explain 35.98% of the variance. 

In order to fully understand the transmission type impact considering other variables, we need to create a multivariate model.

###Building and selecting the Model

To start, we build the first model which considers all variables as predictors. Then we proceed to select the most significant predictors to build the best model.

The step function performs the best model selection, it calls lm repeatedly to build regression models selecting the most significant variables that can be considered as relevant mpg predictors, while discarding the less significant ones.
```{r}
firstmodel <- lm(mpg ~ ., data=mtcars)
bestmodel <- step(firstmodel, direction="both", trace=0)
summary(bestmodel)
anova(firstmodel,bestmodel)
```
By displaying the model summary, we can see that Weight and Accelaration are also relevant to predict mpg. Here the Multiple R-squared value of .8497 means the model can now explain 84.97% of the mpg variance, making it a better predictive model. We also computed the analysis of variance (or deviance) tables for one the previously fitted models (firstmodel and bestmodel).

###Analysis of residuals

Now with the model selection done, we proceed with the analysis of residuals. While reading the following observations, please see **Figure 3 - Residual plots.**

**Observations**

**Residual vs Fitted plot**
The independence condition may be confirmed with the randomly scatterd points that this plot shows.

**Normal Q-Q plot**
Normal distribution of residuals can be confirmed as most points fall in the line.

**Scale-Location plot**
Constant variance can be observed as points are scattered in a constant band.

**Residuals vs Leverage plot**
In this plot we can observe some outliers or leverage points, the points in the top right may indicate values with increased leverage of outliers. We will also perform of regression diagnostics using the function hatvalues() to find data points with most leverage and dfbetas() to find the data points that have bigger influence in the model coefficients.

```{r}
leverage <- hatvalues(bestmodel)
tail(sort(leverage),3)
influential <- dfbetas(bestmodel)
tail(sort(influential[,4]),3)
```

##Conclusions

- MT Vehicles are better to get more MPG compared with AT vehicles.
- MT vehicles have in average 7.24 MPG more than AT vehicles.
- Transmission Type has relevance as a predictor for MPG result, but there are other variables like Weight and Acceleration (wt and qseq) that also have strong influence predicting the mpg outcome.
- The model generated in this analysis has an 85% of accuracy adjusted to the above mentioned variables. 

###Apendix
```{r smallplot, fig.width=3, fig.height=2.5}
boxplot(mpg ~ am, data=mtcars, ylab="Miles per Gallon", xlab="Transmission Type (0=AT, 1=MT)", col=(c("blue","green")))
```

**Figure 1. MPG by Transmission Type - Box plot**

```{r fig.width=6, fig.height=4.5}
require(stats); require(graphics)
pairs(mtcars, pane=panel.smooth, main="Motor Trend Car Road Tests", col=3)
```

**Figure 2. Motor Trend Car Road Tests - Variable pairs plot**

```{r fig.width=6, fig.height=3.3}
par(mfrow=c(2, 2))
plot(bestmodel)
```

**Figure 3. Residual Plots**